Competing Bandits: \\ The Perils of Exploration under Competition

### 2018-11-06 paper/content/introduction.tex
Mostly, edits in related work.

\gaedit{We ask whether algorithms that are traditionally considered ``better" in the standard multi-armed bandit problem are incentivized to be deployed under competition.} We investigate this issue via extensive numerical experiments in a stylized duopoly model. In our model, two firms compete for users and simultaneously learn from them. Each firm commits to a multi-armed bandit algorithm, and \emph{explores} according to this algorithm. Users select between the two firms based on the current reputation score: rewards from the firm's algorithm, averaged over a recent time window. Each firm's objective is to maximize its \gaedit{expected} market share (the fraction of users choosing this firm).

@@

\subsubsection{Related work.}
Our work is related to a longstanding economics literature on competition vs. innovation, \eg \cite{Schumpeter-42,barro2004economic,Aghion-QJE05}. While this literature focuses on R\&D costs of innovation, ``reputational costs" seem new and specific to exploration.

Multi-armed bandits (MAB) is a tractable abstraction for the tradeoff between exploration and \emph{exploitation} (making good near-term decisions based on available information). MAB problems have been studied for many decades, see \cite{Bubeck-survey12} for background. We consider i.i.d. rewards, a well-studied and well-understood MAB model \cite{bandits-ucb1}. We focus on a well-known distinction between ``greedy" (exploitation-only) algorithms, ``naive" algorithms that separate exploration and exploitation, and ``smart" algorithms that combine them. Switching from ``greedy" to ``naive" to ``smart" algorithms involves substantial adoption costs in infrastructure and personnel training \cite{MWT-WhitePaper-2016,DS-arxiv}.

@@

The interplay between exploration, exploitation and incentives has been studied in other scenarios: incentivizing exploration in a recommendation system,
    \eg \cite{Kremer-JPE14,Frazier-ec14,Che-13,ICexploration-ec15,Bimpikis-exploration-ms17},
dynamic auctions
    (see \cite{DynAuctions-survey10} for background),
online ad auctions, \eg
    \cite{MechMAB-ec09,DevanurK09,NSV08,Transform-ec10-jacm,Amin-auctions-nips13},
and human computation
    \cite{RepeatedPA-ec14,Ghosh-itcs13,Krause-www13}.
Our setting is also closely related to the ``dueling algorithms" framework \cite{DuelingAlgs-stoc11}, but this framework considers offline / full feedback scenarios whereas we focus on online machine learning problems.
\swedit{There is also a line of work on ``dueling bandits" (e.g., \cite{Yue-dueling-icml09, Yue-dueling12}) that also studies a form of competition with MAB problems. However, their framework is fundamentally different from ours in that they study learning from setting up ``duels'' between pairs of arms, and we study how learning algorithms compete with each other for data.
}

The study of whether data can give a significant \swdelete{competitive} advantage
has been discussed theoretically in \cite{varian2018artificial,
  lambrecht2015can} and empirically in
\cite{bajari2018impact}. \swedit{While they argue that small amounts
  of additional data do not provide significant improvement in
  training a more predictive model, this set of results focus on learning in isolation, which overlooks certain aspects in a competition.
\swdelete{
These studies consider the effect of additional data in terms of helping towards solving a machine learning problem in isolation and argue that it does not play a large role.} We ask how additional data helps while firms simultaneously solve a machine learning problem and compete for consumers. In our setting, the competition dynamics from our model lead to different conclusions about the competitive advantage that a data advantage can give a firm. Additionally, the first-mover advantage has been well-studied in other settings in economics and marketing \cite{kerin1992first}. }

>>>>

\gaedit{We ask whether algorithms that are traditionally considered ``better" in the literature on multi-armed bandits are incentivized to be deployed under competition.} We investigate this issue via extensive numerical experiments in a stylized duopoly model. In our model, two firms compete for users and simultaneously learn from them. Each firm commits to a multi-armed bandit algorithm, and \emph{explores} according to this algorithm. Users select between the two firms based on the current reputation score: rewards from the firm's algorithm, averaged over a recent time window. Each firm's objective is to maximize its \gaedit{expected} market share (the fraction of users choosing this firm).

@@

\subsection{Related work}

\xhdr{Machine learning.} Multi-armed bandits (MAB) is a tractable abstraction for the tradeoff between exploration and \emph{exploitation} (making good near-term decisions based on available information). MAB problems have been studied for many decades, see \cite{Bubeck-survey12} for background. We consider i.i.d. rewards, a well-studied and well-understood MAB model \cite{bandits-ucb1}. We focus on a well-known distinction between ``greedy" (exploitation-only) algorithms, ``naive" algorithms that separate exploration and exploitation, and ``smart" algorithms that combine them. Switching from ``greedy" to ``naive" to ``smart" algorithms involves substantial adoption costs in infrastructure and personnel training \cite{MWT-WhitePaper-2016,DS-arxiv}.

@@

The interplay between exploration, exploitation and incentives has been studied in other scenarios: incentivizing exploration in a recommendation system,
    \eg \cite{Kremer-JPE14,Frazier-ec14,Che-13,ICexploration-ec15,Bimpikis-exploration-ms17},
dynamic auctions
    (see \cite{DynAuctions-survey10} for background),
online ad auctions, \eg
    \cite{MechMAB-ec09,DevanurK09,NSV08,Transform-ec10-jacm,Amin-auctions-nips13},
and human computation
    \cite{RepeatedPA-ec14,Ghosh-itcs13,Krause-www13}.

Our setting is also closely related to the ``dueling algorithms" framework \cite{DuelingAlgs-stoc11}, but this framework considers offline / full feedback scenarios whereas we focus on online machine learning problems.

\asedit{In ``dueling bandits" (e.g., \cite{Yue-dueling-icml09, Yue-dueling12}), an algorithm sets up a ``duel" between a pair of arms in each round, and only learns which arm has ``won". While this setting features a form of competition inside an MAB problem, it is very different from ours.}

\xhdr{Economics.}
Our work is related to a longstanding economics literature on competition vs. innovation, \eg \cite{Schumpeter-42,barro2004economic,Aghion-QJE05}. While this literature focuses on R\&D costs of innovation, ``reputational costs" seem new and specific to exploration.

\asedit{Whether data gives competitive advantage
has been studied theoretically in \cite{varian2018artificial,
  lambrecht2015can} and empirically in
\cite{bajari2018impact}. 
While these papers find that small amounts
  of additional data do not provide significant improvement,
 % in training a more predictive model, 
  they focus on learning in isolation.
  %, which overlooks certain aspects in a competition.
%\swdelete{These studies consider the effect of additional data in terms of helping towards solving a machine learning problem in isolation and argue that it does not play a large role.} 
We ask how additional data helps when firms simultaneously solve a machine learning problem and compete for consumers. Due to the competition dynamics in our model, we find that even a small amount of additional data tends to help a lot.}

The first-mover advantage has been well-studied in other settings in economics and marketing, see survey \cite{kerin1992first}.
